A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-13 Simulation/Writeup/NoteConvergence.tex
results

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
It. & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}

\textbf{$\widehat{R}$}. The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct! Also say something about Rh below 1.}

\textbf{autocorrelation.} All observed autocorrelations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %-0.735 
at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data. \textbf{However, using the 1.2 threshold may be to lenient because this was observed at three iterations and then again after six. If we would use 1.2, we might miss this increase after three iterations. The 1.1 threshold seems OK. Also, ten iterations are computationally not terrible. With increased performance and storage capacities, doubling the default is fine I guess.}

@@

An implication of the observed dip in AC is that default maxit value of five iterations is the worst possible number of iterations.

\textbf{Future research.} This study considered only a MCAR missingness mechanism. Necessary but not sufficient to demonstrate the performance of convergence diagnostics. This is just a proof of concept. Further research is needed to investigate the performance under violation of convergence, e.g. dependency between predictors (very high correlations). %We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}

>>>>

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
T & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$). Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero). The convergence diagnostics show more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.

%The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
\textbf{say something about Rh below 1.}
%%%%

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. At $T=5$, we reach the same $AC$-value as observed at $T=2$. $AC$-values plateau around $T=70$. $AC$ is then indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

% All observed autocorrelations  are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %-0.735 
% at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data. \textbf{However, using the 1.2 threshold may be too lenient because this was observed at three iterations and then again after six. If we would use 1.2, we might miss this increase after three iterations. The 1.1 threshold seems OK. Also, ten iterations are computationally not terrible. With increased performance and storage capacities, doubling the default is fine I guess.}

@@

The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.

\textbf{Future research.} This study considered only an MCAR missingness mechanism. Necessary but not sufficient to demonstrate the performance of convergence diagnostics. This is just a proof of concept. Further research is needed to investigate the performance under violation of convergence, e.g. dependency between predictors (very high correlations). %We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}
